---
title: "Hands-On Tutorials for Covariate Adjustment in Randomized Trials"
subtitle: "Binary Outcomes: Standardization (G-Computation)"
author: "Josh Betz (jbetz@jhu.edu), Kelly Van Lancker (kvanlan3@jhu.edu), and Michael Rosenblum (mrosen@jhu.edu)"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
header-includes:
   - \usepackage{amsmath}
output:
  html_document:
    self_contained: TRUE
    toc: TRUE
    toc_float: TRUE
    toc_depth: 2
    number_sections: TRUE  ## if you want number sections at each table header
    theme: united
    highlight: tango
    code_folding: show
---

```{r Check-Packages, include = FALSE, echo = FALSE, message = FALSE}
parameters_report <-
  config::get(
    config = "report_parameters",
    file = file.path(here::here(), "config.yml")
  )

parameters_example <-
  config::get(
    config = "binary_example_mistie",
    file = file.path(here::here(), "config.yml")
  )

installed_packages <- installed.packages()[, "Package"]

required_packages <-
  with(
    data = parameters_report,
    c(cran_packages,
      gsub(
        pattern = "^[A-Za-z0-9\\_\\.\\-]*/",
        replacement = "",
        x = github_packages
      )
    )
  )

missing_packages <-
  setdiff(
    x = required_packages,
    y = installed_packages
  )

if(length(missing_packages) > 0){
  stop(
    "Required packages not installed: ", 
    paste(missing_packages, collapse = ","), ". ",
    "See `cran_packages` and `github_packages` in the configuration file."
  )
}

example_output_path <-
  file.path(here::here(), parameters_example$example_output_file_name)

if(!file.exists(example_output_path)){
  stop(
    "Example data does not exist."
  )
} else {
  temp_environment <- new.env()
  
  load(
    file = example_output_path,
    envir = temp_environment
  )
  
  rename_params <-
    c("parameters_report")
  
  import_objects <-
    setdiff(
      x = names(temp_environment),
      y = rename_params
    )
  
  for(i in 1:length(import_objects)){
    assign(
      x = import_objects[i],
      value = get(x = import_objects[i], envir = temp_environment)
    )
  }
  
  rm(temp_environment)
}
```



## Report Setup {.tabset}

### Load Packages

```{r Load-Packages}
library(cobalt) # Covariate Balance
library(dplyr) # Wrangling Results
library(impart) # Carrying Out Analyses
```


```{r Report-Setup, include = FALSE, echo = FALSE, message = FALSE}
with(
  data = 
    parameters_report$opts_chunk,
  expr = {
    knitr::opts_chunk$set(
      collapse = collapse_handout,
      echo = echo_handout,
      message = message,
      error = error,
      purl = purl,
      results = results_handout,
      fig.path = paste0(fig_path),
      fig.width = fig_width,
      fig.height = fig_height,
      fig.align = fig_align,
      out.width = out_width,
      dpi = dpi
    )
  }
)
```


## Study Descriptives {.tabset}

### Covariates: By Treatment, Aggregate {.smaller}

```{r Covariate-Summaries}
table1::table1(
  x = ~ 
    age + male + hx_cvd + hx_hyperlipidemia +
    on_anticoagulants + on_antiplatelets +
    ich_location + ich_s_volume + ivh_s_volume + 
    gcs_category |
    arm,
  data = sim_miii_fixed
)
```




### Covariate Balance

```{r Covariate-Balance}
cobalt::bal.tab(
  x = 
    sim_miii_fixed %>% 
    dplyr::select(
      age, male, on_antiplatelets, ich_location, ich_s_volume,
      ivh_s_volume, gcs_category
    ),
  treat = sim_miii_fixed$arm,
  binary = "std",
  continuous = "std",
  s.d.denom = "pooled"
)
```




### Outcomes:

```{r Table-Outcomes}
table1::table1(
  x = ~ 
    mrs_365d +
    mrs_365d_binary_factor |
    arm,
  data = sim_miii_fixed
)
```




### Crude Estimates

```{r Crude-Estimates}
y1_hat <-
  with(
    data = subset(sim_miii_fixed, arm == "surgical"),
    mean(mrs_365d_binary, na.rm = TRUE)
  )
y0_hat <-
  with(
    data = subset(sim_miii_fixed, arm == "medical"),
    mean(mrs_365d_binary, na.rm = TRUE)
  )

# Risk Difference
y1_hat - y0_hat

# Risk Ratio
y1_hat/y0_hat

# Odds Ratio
(y1_hat*(1 - y0_hat))/(y0_hat*(1 - y1_hat))
```




--------------------------------------------------------------------------------




## Unadjusted Analysis {.tabset}

### Fitting Logistic Model

```{r Fit-Unadjusted-Model}
mrs_unadjusted_logistic_glm <-
  stats::glm(
    formula = mrs_365d_binary ~ arm,
    data = sim_miii_fixed,
    family = binomial(link = "logit")
  )
summary(mrs_unadjusted_logistic_glm)
```




### Generating Unadjusted Predictions

```{r Generate-Unadjusted-Predictions}
pr_outcome_unadj_control <-
  stats::predict(
    object = mrs_unadjusted_logistic_glm,
    newdata = 
      within(data = sim_miii_fixed,
             expr = {arm = "medical"}),
    type = "response"
  )

pr_outcome_unadj_treatment <-
  stats::predict(
    object = mrs_unadjusted_logistic_glm,
    newdata = 
      within(data = sim_miii_fixed,
             expr = {arm = "surgical"}),
    type = "response"
  )

table(pr_outcome_unadj_control)
table(pr_outcome_unadj_treatment)
```



### Average & Contrast Predictions

```{r Contrast-Average-Unadjusted-Predictions 1}
e_y_0_unadj <- mean(pr_outcome_unadj_control)
e_y_1_unadj <- mean(pr_outcome_unadj_treatment)

# Risk Difference
e_y_1_unadj - e_y_0_unadj
# Relative Risk
e_y_1_unadj/e_y_0_unadj
# Odds Ratio
(e_y_1_unadj*(1 - e_y_0_unadj))/
  (e_y_0_unadj*(1 - e_y_1_unadj))
```




### Compare with Logistic GLM

```{r Compute-Unadjusted-Contrasts-from-Model}
unadj_glm_beta <- coef(mrs_unadjusted_logistic_glm)
pr_medical <-
  plogis(unadj_glm_beta["(Intercept)"])

pr_surgical <-
  plogis(unadj_glm_beta["(Intercept)"] +
           unadj_glm_beta["armsurgical"])

pr_surgical - pr_medical # Risk Difference
pr_surgical/pr_medical # Relative Risk
exp(unadj_glm_beta["armsurgical"]) # Odds Ratio
```




### Worked Example: Unadjusted

```{r Compute-Unadjusted-ATE}
source("boot_p_value.R")
library(boot)

set.seed(12345) # Set seed for Reproducibility

rd_unadjusted_result <-
  impart::standardization(
    data = sim_miii_fixed,
    outcome_formula = mrs_365d_binary ~ tx,
    family = binomial(link = "logit"),
    estimand = "difference", # "ratio" or "oddsratio"
    treatment_column = "tx",
    se_method = "bootstrap", # "none" "influence" "score"
    verbose = TRUE # Needed for p-value, not CI
  )

with(
  data = rd_unadjusted_result$result,
  expr = {
    c(
      "Estimate" = estimate,
      "SE" = se_boot_df_adjusted,
      "LCL" = lcl_bca_df_adjusted_0.95,
      "UCL" = ucl_bca_df_adjusted_0.95
    )
  }
)

boot_p_value(
  boot_object = rd_unadjusted_result$boot_object,
  ci_method = "bca",
  null_value = 0,
  # Adjust CI width for DF in model
  var_adjust = rd_unadjusted_result$result$variance_factor
)
```




--------------------------------------------------------------------------------




## Covariate Adjusted Analyses {.tabset}


### Fit Adjusted Logit GLM

```{r Fit-Adjusted-Model-1, results = "markup"}
mrs_adjusted_logistic_glm <-
  stats::glm(
    formula = 
      mrs_365d_binary ~ arm +
      age + 
      male +
      hx_cvd +
      hx_hyperlipidemia +
      on_anticoagulants +
      on_antiplatelets +
      ich_location +
      ich_s_volume +
      ivh_s_volume + 
      gcs_category,
    data = sim_miii_fixed,
    family =
      binomial(link = "logit")
  )

summary(mrs_adjusted_logistic_glm)
```




### Generating Adjusted Predictions

```{r Generate-Adjusted-Predictions}
pr_outcome_adj_control <-
  stats::predict(
    object = mrs_adjusted_logistic_glm,
    newdata = 
      within(data = sim_miii_fixed,
             expr = {arm = "medical"}),
    type = "response"
  )

pr_outcome_adj_treatment <-
  stats::predict(
    object = mrs_adjusted_logistic_glm,
    newdata = 
      within(data = sim_miii_fixed,
             expr = {arm = "surgical"}),
    type = "response"
  )

par(mfrow = c(1, 2))
hist(
  pr_outcome_adj_treatment,
  breaks = "fd"
)

hist(
  pr_outcome_adj_control,
  breaks = "fd"
)
```




### Results: Adjusted

```{r Compute-Adjusted-ATE, results = "markup"}
rd_adjusted_result <-
  impart::standardization(
    data = sim_miii_fixed,
    outcome_formula = mrs_365d_binary ~ tx + age + male + hx_cvd +
      hx_hyperlipidemia + on_anticoagulants + on_antiplatelets + ich_location +
      ich_s_volume + ivh_s_volume + gcs_category,
    family = binomial(link = "logit"),
    estimand = "difference", # "ratio" or "oddsratio"
    treatment_column = "tx",
    se_method = "bootstrap", # "none" "influence" "score"
    verbose = TRUE # Needed for p-value, not CI
  )

with(
  data = rd_adjusted_result$result,
  expr = {
    c(
      "Estimate" = estimate,
      "SE" = se_boot_df_adjusted,
      "LCL" = lcl_bca_df_adjusted_0.95,
      "UCL" = ucl_bca_df_adjusted_0.95
    )
  }
)

boot_p_value(
  boot_object = rd_adjusted_result$boot_object,
  ci_method = "bca",
  null_value = 0,
  # Adjust CI width for DF in model
  var_adjust = rd_adjusted_result$result$variance_factor
)
```




--------------------------------------------------------------------------------

## Influence Function vs. Bootstrap

```{r Compute-ATE-Influence, results = "markup"}
impart::standardization(
  data = sim_miii_fixed,
  outcome_formula = mrs_365d_binary ~ tx,
  family = binomial(link = "logit"),
  estimand = "difference", # "ratio" or "oddsratio"
  treatment_column = "tx",
  se_method = "influence", # "none" "influence" "score"
)

impart::standardization(
  data = sim_miii_fixed,
  outcome_formula = mrs_365d_binary ~ tx + age + male + hx_cvd +
    hx_hyperlipidemia + on_anticoagulants + on_antiplatelets + ich_location +
    ich_s_volume + ivh_s_volume + gcs_category,
  family = binomial(link = "logit"),
  estimand = "difference", # "ratio" or "oddsratio"
  treatment_column = "tx",
  se_method = "influence", # "none" "influence" "score"
)
```




--------------------------------------------------------------------------------

## Comparing Analysis Methods:

  - $Var(\theta) = (SE(\theta))^2$
  - Relative Efficiency: $RE_{A/U} = Var(\theta_{U})/Var(\theta_{A})$
  - Relative Change in Variance: $RCV_{A/U} = \frac{Var(\theta_{A}) - Var(\theta_{U})}{Var(\theta_{U})} = \frac{1}{RE_{A/U}} - 1$
  - Relative Change in Precision: $RCP_{A/U} = \frac{1/Var(\theta_{A}) - 1/Var(\theta_{U})}{1/Var(\theta_{U})} = Var(\theta_{U})/Var(\theta_{A}) - 1 = RE_{A/U} - 1$
